Early warning signals: The charted and uncharted territories

### 2013-05-29 arxiv-copy.tex
building arxiv copy of pdf, slightly hand-tuned pdf * adjusted pandoc's longtable environment to include \onecolumn and appear as a float * pdflatex flavor instead of xelatex (or vanilla latex flavor) * see make fancy

\documentclass{article} 
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.

\title{Early warning signals: The charted and uncharted territories}
\author{%
  Carl Boettiger\thanks{CB \& NR contributed equally. Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA} \and Noam Ross\thanks{Department of Environmental Science and Policy, University of California Davis, 1 Shields Avenue, Davis, CA 95616 USA} \and Alan Hastings\footnotemark[2]
}
\begin{document}

\maketitle

@@

The realization that complex systems such as ecological communities can collapse or shift regimes suddenly and without rapid external forcing poses a serious challenge to our understanding and management of the natural world.  The potential to identify early warning signals that would allow researchers and managers to predict such events before they happen has therefore been an invaluable discovery that offers a way forward in spite of such seemingly unpredictable behavior.  Research into early warning signals has demonstrated that it is possible to define and detect such early warning signals in advance of a transition in certain contexts.  Here we describe the pattern emerging as research continues to explore just how far we can generalize these results.  A core of examples emerges that shares three properties: the phenomenon of rapid regime shifts,  a pattern of 'critical slowing down' that can be used to detect the approaching shift, and a mechanism of bifurcation driving the sudden change.  As research has expanded beyond these core examples, it is becoming clear that not all systems that show regime shifts exhibit critical slowing down, or vice versa.  Even when systems exhibit critical slowing down, statistical detection is a challenge.  We review the literature that explores these edge cases and highlight the need for (a) new early warning behaviors that can be used in cases where rapid shifts do not exhibit critical slowing down, (b) the development of methods to identify which behavior might be an appropriate signal when encountering a novel system; bearing in mind that a positive indication for some systems is a negative indication in others, and (c) statistical methods that can distinguish between signatures of early warning behaviors and noise.
\\
Key words: early warning signals, regime shifts, bifurcation, critical slowing down 
\\
\end{abstract}
%\footnote[*]{authors contributed equally}
%\footnote[a]{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA}
%\footnote[b]{Department of Environmental Science and Policy, University of California Davis, 1 Shields Avenue, Davis, CA 95616 USA}

@@

\begin{longtable}[c]{@{}ll@{}}
\hline\noalign{\medskip}
\begin{minipage}[t]{0.42\columnwidth}\raggedright
\textbf{Critical slowing down (CSD)}
\end{minipage} & \begin{minipage}[t]{0.55\columnwidth}\raggedright
A system's slowing response to perturbations as it's dominant eigenvalue
approaches zero, often expressed in greater variance, autocorrelation,
and return time. CSD is one possible EWS.
\end{minipage}
\\\noalign{\medskip}
\begin{minipage}[t]{0.42\columnwidth}\raggedright
\textbf{Early warning signals (EWS)}
\end{minipage} & \begin{minipage}[t]{0.55\columnwidth}\raggedright
A general term for dynamic patterns in system behavior that precede
regime shifts. Though CSD phenomena are among the best studied EWS, some
shifts will require alternative signals; Figure 1.
\end{minipage}
\\\noalign{\medskip}
\hline
\end{longtable}

\textbf{Definitions}: In this paper we refer to two closely related but
different phenomena

@@

\begin{figure}[htbp]
\centering
\includegraphics{ews-venn.eps}
\caption{Venn diagram representing the intersecting domains of rapid
regime shifts, bifurcations, and critical slowing down. Labels in italic
are example phenomena that occur in each domain. Roman numerals and
indicate example literature (right) exploring each domain, and also
refer to sections below describing those domains. Each dot represents a
study in the domain. Studies and dots in grey represent literature not
explicitly testing EWS, but which demonstrate phenomena related to EWS.
The center domain (I) where all three phenomena intersect, is the most
extensively researched domain of the EWS field. Literature outside this
charted region does not yet provide the needed EWS, but hints where
existing signals based on CSD may be insufficient or misleading.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{schreiber-fig.eps}
\caption{A system where variance decreases prior to a population
collapse; adapted from Schreiber (2003). In this model, prey species
with high growth rates exhibit chaotic dynamics under predation, but
populations collapse when predation increases beyond a threshold value.
Left: The population level as a function of predation rate. Mean
dynamics shown as black line, realizations with varying initial
conditions shown as grey dots; see Schreiber (2003). Middle: Variance of
the prey population level. Note that it \emph{decreases} as predation
rate approaches the threshold. Right: Lag-1 Autocorrelation in prey
population dynamics increases as the threshold is approached}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{kefi-fig.eps}
\caption{A system where critical slowing down is observed without a
critical threshold, from Kéfi et al. (2012). In this model, prey have
logistic growth and are subject to predation with a Type III functional
response, but there is no bifurcation. Instead, average prey population
exhibits a smooth response to increased predation (grazing). Left: The
population level as a function of predation rate. Middle: Variance of
the prey population level. Right: Lag-1 Autocorrelation in prey
population dynamics as grazing rate increases. Note that both indicators
increase despite the lack of a bifurcation.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{Bel2012example.eps}
\caption{Difference between different types of perturbations. On the
horizontal axis is the bifurcation parameter, representing the state of
the environment (e.g.~annual mean temperature) whose slow change could
lead to a sudden shift. A direct disturbance to the system state
(e.g.~population size, vertical axis) could also cause a transition if
it is large enough to cross the stability threshold (dashed line). Such
a perturbation can come from exogenous factors such as anthropogenic
pressures or occur by chance from intrinsic stochasticity. These
distinct mechanisms of disturbance and environmental change are coupled
-- as the environment deteriorates, moving the system right on the
diagram, the probability that a disturbance crosses the threshold
increases. From Bel et al. (2012).}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{ROC.eps}
\caption{Receiver-operating characteristic (ROC) curves illustrate the
trade-off between false positive and true positive detection rates of an
EWS. Perfect warning signals (solid curve) would identify all thresholds
while generating no false positives, while very poor signals would have
no ability to distinguish false from true signals (dotted line). In
reality, warning signals' have a trade-off between the two which is
described by a curve (dotted line) or summarized by the area under the
ROC curve}
\end{figure}

>>>>

\documentclass[author-year, 5p]{elsarticle} %review=doublespace preprint=single 5p=2 column
%\documentclass[author-year, review]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add 
%\linenumbers % turns line numbering on 
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

% A modified page layout
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
  \usepackage[unicode=true]{hyperref}
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Early warning signals: The charted and uncharted territories},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\begin{document}
\begin{frontmatter}
  \title{Early warning signals: The charted and uncharted territories}
  \author[cstar]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@gmail.com}
  \author[esp]{Noam Ross\corref{cor1}}
  \author[esp]{Alan Hastings}
  \cortext[cor1]{authors contributed equally}
  \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA}
  \address[esp]{Department of Environmental Science and Policy, University of California Davis, 1 Shields Avenue, Davis, CA 95616 USA}

@@

The realization that complex systems such as ecological communities can collapse or shift regimes suddenly and without rapid external forcing poses a serious challenge to our understanding and management of the natural world.  The potential to identify early warning signals that would allow researchers and managers to predict such events before they happen has therefore been an invaluable discovery that offers a way forward in spite of such seemingly unpredictable behavior.  Research into early warning signals has demonstrated that it is possible to define and detect such early warning signals in advance of a transition in certain contexts.  Here we describe the pattern emerging as research continues to explore just how far we can generalize these results.  A core of examples emerges that shares three properties: the phenomenon of rapid regime shifts,  a pattern of 'critical slowing down' that can be used to detect the approaching shift, and a mechanism of bifurcation driving the sudden change.  As research has expanded beyond these core examples, it is becoming clear that not all systems that show regime shifts exhibit critical slowing down, or vice versa.  Even when systems exhibit critical slowing down, statistical detection is a challenge.  We review the literature that explores these edge cases and highlight the need for (a) new early warning behaviors that can be used in cases where rapid shifts do not exhibit critical slowing down, (b) the development of methods to identify which behavior might be an appropriate signal when encountering a novel system; bearing in mind that a positive indication for some systems is a negative indication in others, and (c) statistical methods that can distinguish between signatures of early warning behaviors and noise.
  \end{abstract}
  \begin{keyword}
early warning signals \sep regime shifts \sep bifurcation \sep critical slowing down 
   \end{keyword}

 \end{frontmatter}

@@

\begin{figure}[htbp]
\onecolumn
\begin{longtable}[l]{@{}ll@{}}
\hline\noalign{\medskip}
\begin{minipage}[t]{0.10\columnwidth}\raggedright
\textbf{Critical slowing down (CSD)}
\end{minipage} & \begin{minipage}[t]{0.35\columnwidth}\raggedright
A system's slowing response to perturbations as it's dominant eigenvalue
approaches zero, often expressed in greater variance, autocorrelation,
and return time. CSD is one possible EWS.
\end{minipage}
\\\noalign{\medskip}
\begin{minipage}[t]{0.10\columnwidth}\raggedright
\textbf{Early warning signals (EWS)}
\end{minipage} & \begin{minipage}[t]{0.35\columnwidth}\raggedright
A general term for dynamic patterns in system behavior that precede
regime shifts. Though CSD phenomena are among the best studied EWS, some
shifts will require alternative signals; Figure 1.
\end{minipage}
\\\noalign{\smallskip}
\hline
\end{longtable}
\twocolumn
\textbf{Definitions}: In this paper we refer to two closely related but
different phenomena
\end{figure}

@@

\begin{figure*}[htbp]
\centering
\includegraphics{ews-venn.pdf}
\caption{Venn diagram representing the intersecting domains of rapid
regime shifts, bifurcations, and critical slowing down. Labels in italic
are example phenomena that occur in each domain. Roman numerals and
indicate example literature (right) exploring each domain, and also
refer to sections below describing those domains. Each dot represents a
study in the domain. Studies and dots in grey represent literature not
explicitly testing EWS, but which demonstrate phenomena related to EWS.
The center domain (I) where all three phenomena intersect, is the most
extensively researched domain of the EWS field. Literature outside this
charted region does not yet provide the needed EWS, but hints where
existing signals based on CSD may be insufficient or misleading.}
\end{figure*}

@@

\begin{figure}[htbp]
\centering
\includegraphics{schreiber-fig.pdf}
\caption{A system where variance decreases prior to a population
collapse; adapted from Schreiber (2003). In this model, prey species
with high growth rates exhibit chaotic dynamics under predation, but
populations collapse when predation increases beyond a threshold value.
Left: The population level as a function of predation rate. Mean
dynamics shown as black line, realizations with varying initial
conditions shown as grey dots; see Schreiber (2003). Middle: Variance of
the prey population level. Note that it \emph{decreases} as predation
rate approaches the threshold. Right: Lag-1 Autocorrelation in prey
population dynamics increases as the threshold is approached}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{kefi-fig.pdf}
\caption{A system where critical slowing down is observed without a
critical threshold, from Kéfi et al. (2012). In this model, prey have
logistic growth and are subject to predation with a Type III functional
response, but there is no bifurcation. Instead, average prey population
exhibits a smooth response to increased predation (grazing). Left: The
population level as a function of predation rate. Middle: Variance of
the prey population level. Right: Lag-1 Autocorrelation in prey
population dynamics as grazing rate increases. Note that both indicators
increase despite the lack of a bifurcation.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{Bel2012example.pdf}
\caption{Difference between different types of perturbations. On the
horizontal axis is the bifurcation parameter, representing the state of
the environment (e.g.~annual mean temperature) whose slow change could
lead to a sudden shift. A direct disturbance to the system state
(e.g.~population size, vertical axis) could also cause a transition if
it is large enough to cross the stability threshold (dashed line). Such
a perturbation can come from exogenous factors such as anthropogenic
pressures or occur by chance from intrinsic stochasticity. These
distinct mechanisms of disturbance and environmental change are coupled
-- as the environment deteriorates, moving the system right on the
diagram, the probability that a disturbance crosses the threshold
increases. From Bel et al. (2012).}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{ROC.pdf}
\caption{Receiver-operating characteristic (ROC) curves illustrate the
trade-off between false positive and true positive detection rates of an
EWS. Perfect warning signals (solid curve) would identify all thresholds
while generating no false positives, while very poor signals would have
no ability to distinguish false from true signals (dotted line). In
reality, warning signals' have a trade-off between the two which is
described by a curve (dotted line) or summarized by the area under the
ROC curve}
\end{figure}
